---
title: "Planning for the pollock assessment in the EBS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Planning for the pollock assessment in the EBS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ebswp)
```

# EBS 2023 assessment analyses planning

## From the 2022 SSC minutes:

**The SSC suggests that walleye pollock is a good candidate for
considering the impacts of highly variable recruitment on reference
points in the context of the Council's harvest control rules** (see
discussion on working groups in the JGPT report section). For example,
the SAFE authors suggested exploring an explicit harvest control rule
that maintains productivity at the level observed over recent decades
(p. 33). The SSC supports considerations of modified harvest control
rules, particularly for stocks with highly variable and uncertain
recruitment. If the Council chooses, this could include considerations
for stabilizing catches over time or including other economic
considerations in the harvest control rules.

-   *For the 2023 assessment we examine the variability of the biological reference points historically and note that there is general stability in the 
$B_{MSY}$ 
estimates. The Tier 1 ABC/OFL cthealculations can result in highly variable 
estimates as the stock approaches
$B_{MSY}$ and drops below that value (as happend in the 2009-2010 period).*


The SSC had the following additional recommendations for the authors:

-   Maturity and growth information from the NBS has not been examined
    yet. Given the possible importance of the NBS to walleye pollock and
    other species in the future, the SSC suggests this should be a high
    priority.

    -   *These data being processed and this work  is underway*

-   The SSC supports efforts to implement recent advances in improving
    the statistical treatment of compositional data using the Dirichlet
    distribution or other approaches.

    -   *Tradeoffs in data weighting were pursued in September 2023 and
        saught to find a balance between observation error and process
        errors*

-   The SAFE document lists a number of research recommendations (p.
    36/37). The SSC notes that some of these are at least in progress.
    The SSC generally supports these recommendations but requests that
    the authors update the list of priorities to clarify to what extent
    some of these priorities have been partially or fully addressed.

    -   *We updated the priorities and listed those that have been
        completed or are continued to be underway*"

-   In particular, the SSC notes that genetic sample collection and
    analyses are listed as a research priority across all pollock stocks
    and that some work has been completed. The SSC highlights the
    importance of additional genetics work and would appreciate an
    update on the status of this work either as part of the assessment
    or separately.

    -   *We revisited the stock structure work attached as an appendix to the 2015 
    SAFE report chapter and are examining the extent that this work
    needs updating.
        *"

-   The SSC appreciated the adjustments to weight-at-age in the survey
    that was included in this year's assessment and suggests that these
    changes may be substantial enough to warrant an examination of their
    impact on assessment results.

    -   *This publication has been completed and in the present assessment
    we evaluated the implication of alternative spawning biomass-at-age 
    assumptions.*"

-   With respect to the multi-species CEATTLE model, the SSC concurs
    with Plan Team recommendations to use the model to inform risk table
    discussions and to consider ways in which model outputs, in
    particular estimates of predation mortality, can inform
    single-species assessments.

    -   *We wil include a more thorough evaluation and discussion in the 
    2023 assessment.*

-   The SSC encourages the authors to consider model-based solutions to
    uncertain recruitment estimates rather than ad-hoc adjustments. In
    particular, reductions in the assumed recruitment variance parameter
    may result in less extreme recruitment estimates. Other systematic
    approaches to addressing the uncertainty may also be considered.

    -   *We revisited applying the age-determination error matrix as 
    a sensitivity as this can impact the recruitment variability and 
    estimation uncertainty.*

-   The SSC suggests that authors include a plot to compare estimates of
    recent recruitments as they change over time similar to Fig 3.33
    (pg. 88) in the sablefish assessment.

    -   *We provide a figure of estimated recruitment by year class
        (1977 -- 2019) in number of age-1 fish (billions of fish) for
        the 2022 and 2023 models.* "

-   The SSC supports the move across assessment from design-based
    estimates of survey biomass to VAST estimates. The SSC recommends
    that the design-based estimates be produced as a check on VAST
    estimates and as a fallback option if needed, although they may not
    need to be included in the assessment.

    -   *We provide a table showing the design-based estimates and conduct
    a model run with those estimates. This may be an approach to adopt so
    that bridging across assessment modeling platforms can be facilitated
    (most other assessment model platforms are unable to deal with index
    time series that have a covariance matrix)*

### Tasks and todo list

  [ ] Add in survey biomass\
  [ ] Update catches\
  [ ] Add in 2022 fishery age compositions\
  [ ] Add in 2023 survey age compositions\
  [x] Modify code to accommodate FT-NIRS and other age-determination error
  matrices by gear and year 
  [ ] Figure out which models to use going
    forward

## From the 2021 SSC minutes

-   Ongoing genetic studies to determine the relationship between
    pollock in the NBS and EBS, and nearby GOA and AI regions.

-   The 2019 BSAI GPT recommendation to revisit and evaluate the
    treatment of variance parameters within the assessment, with
    particular attention to those that are fixed.

-   Efforts to quantify pollock movement and abundance along the
    US-Russia EEZ boundary.

-   Geostatistical analyses of combined trawl and acoustic data to
    provide a single time-series, statistically accounting for the
    overlap between these data, for informing stock trends.

## The SSC provides the following additional recommendations:

-   Exploration of young-of-year pollock density and quality estimates
    from NMFS BASIS surveys to inform pollock recruitment.

-   Consideration of whether the observed sensitivity in the SRR to
    prior specification should constitute an increased risk level
    specification within the assessment or population dynamicsrelated
    considerations. This could provide a clearer justification for the
    use of the Tier 3 calculation as the basis for harvest
    specification.

-   Given the time-varying specification of fishery selectivity within
    the assessment model and the large change in the estimated 2021 FOFL
    between the 2019 and 2020 assessments, the authors should provide a
    retrospective comparison of the selectivity assumed in projections
    to that estimated with the addition of new data.

-   Consideration of whether risk table specifications should account
    for the importance of pollock as a key forage species in the EBS
    ecosystem to better justify the use of a Tier 3 ABC determination as
    a precautionary measure for this Tier 1 stock.

-   Given the apparent disappearance of the second and large mode in
    fishery length compositions as the 2020 B-season progressed,
    exploration of within-season spatial variation in fishery length
    composition would be useful in evaluating whether these larger
    pollock simply moved out of the area of fishing effort, or died as a
    result of natural or fishing mortality.

| Models | 3m  | 0.5m | Standard survey | Dens. Dep. Correction | Numbers | Biomass | Numbers | Biomass | 15.1 Input sample sizes | Revised input sample sizes | wt 15.1 | wt 16.0 | Description                             |
|------|------|------|------|------|------|------|------|------|------|------|------|------|------|
| 14.1   | x   |      | x               |                       | x       |         | x       |         | x                       |                            | x       |         | 2014 model                              |
| 15.1   | x   |      |                 | x                     | x       |         | x       |         | x                       |                            | x       |         | 2015 model                              |
| 16.01  | x   |      | x               |                       |         | x       |         | x       | x                       |                            | x       |         | Transition to biomass                   |
| 16.02  |     | x    | x               |                       |         | x       |         | x       | x                       |                            | x       |         | Transition to ATS 0.5m                  |
| 16.03  |     | x    |                 | x                     |         | x       |         | x       |                         |                            | x       |         | Transition to BTS dens. dep. Correction |
| 16.04  |     | x    |                 | x                     |         | x       |         | x       |                         | x                          | x       |         | Input sample size adjustment            |
| 16.1   |     | x    |                 | x                     |         | x       |         | x       |                         | x                          |         | x       | Proposed model                          |
