---
title: "Planning notes for the pollock assessment in the EBS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Planning notes for the pollock assessment in the EBS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::quarto}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ebswp)
```

# EBS pollock 2024 assessment analyses planning

## From the 2023 SSC minutes:

-  The SSC would prefer not to make a risk table adjustment based on the difference from Tier 1 to Tier 3 again during the 2024 assessment cycle. The SSC requests that the next stock assessment bring back a new approach that may include development of a constant buffer based on factors extrinsic to the stock assessment (ecosystem function), or a better representation of the uncertainty in the Tier 1 and control rule calculations such that a reduction from maximum ABC is not needed every year.
  
    -  *Will approach this with a discussion of alternative HCRs for pollock that consider ecosystem function.  Re-examine what was done in 2011, consider a proper alternative feedback approach, noting that the early work including climate change.*
    
  

-  Use posterior distributions from the MCMC to determine probabilities in the risk table and expand the columns in the risk table to include the recommended ABC (and potentially higher values).
  
      - *Just requires some code modifications to do the computations off of the posteriors*


-  Identify where MLE estimates are being used and where MCMC estimates are being used. Also see the SSC’s General Stock Assessment Comments to include convergence diagnostics any time Bayesian results are reported. If MCMC diagnostics continue to appear adequate, reference points could be calculated using the posterior distribution used, rather than an analytical calculation.
  
     -   *MLE is used and asymptotic approximations are used everywhere for advice, MCMC has been used as a comparison only and to show inter-relationships. Just requires some code modifications to do the computations off of the posteriors*


-  The SSC recommends that consideration be given to removal of the Japanese fishery CPUE index (1965-76) from the assessment, because this data set no longer seems to contribute to the assessment. A sensitivity test should be done to evaluate the effects of data removal on the assessment.

    - *Will investigate. In 1964 data suggests all young small fish from the north, these data are handed down, issues about catching being not represenative of population, doesn't impact much, makes it hard to get different assessment models running so may make sense to remove*


-  Catch-at-age data provided by foreign fishing agencies in the pre-Magnuson era were not produced using the same aging criteria as the AFSC age-and-growth program. Consideration should be given to removal of these data from the assessment. A sensitivity test should be done to evaluate the effects of data removal on the assessment.

    - *Will investigate, age-determination methods from historical records and/or Russian scientists.*


- Document the method used for determining the selectivity to use in the forward projections and continue to evaluate projection variability due to selectivity. The SSC appreciates the selectivity retrospective comparison and suggests that it might be helpful to limit the comparison to the projection used in each year against only the most recent (best) estimate of selectivity for that year.

    - *Better document selectivity assumptions for projections. Look at Cole's approach for plotting out performance. Consider historical uncertainty and variability.*


- The SSC supports the use of posterior predictive distributions, an underutilized tool in fisheries science, but common in other fields. To fully implement this approach to Bayesian model checking the SSC recommends plotting a histogram for each data source of the percentile of the predictive distribution in which each data point lies, noting that in a highly consistent model this histogram would be uniform.

    - *This conclusion might be incorrect. Likely to write that this will evolve into better approaches to include in the future*


-  There is an apparent shift towards older ages in fisheries and trawl survey selectivity that should be investigated further.
 
     - *Will investigate, focused around changes in availabilty.*
     
-  The SSC agrees with the BSAI GPT’s proposal in their presentation to move the multi-species model out of the pollock stock assessment, where it has been included as an appendix since it was first developed. Instead, they suggested it would be a separate chapter listed in parallel with the ESR, as it applies to multiple stocks and informs the ESRs.

    - *Technically it's presented as part of the BSAI assessment. Agree that it should be highlighted on its own.*

-  The SSC suggests revisiting the treatment of the stock-recruit relationship in the assessment model using recent improvements in modeling approaches and a longer time series that encompasses the recent warm period in the EBS. Recruitment deviates should be from the stock-recruit relationship and should model variability among annual recruitment estimates based on information in the data and residual variability. The estimation process should ensure that log-normally distributed recruitments are mean unbiased, resulting in unbiased biomass estimates. If an informative prior is used for steepness, it should be based on a meta-analysis of related species and reflect the uncertainty of that meta-analysis. Further consideration of time periods (as in previous analyses) and the influence of temperature on the stock-recruit relationship may be helpful. The SSC recognizes that there were significant recent analyses in 2016, 2018 and 2020 and is not requesting a repeat of those but a review of previous work would be helpful.

    - *Will include a review and a re-parameterization of numbers at age 1 being from the SRR. We will take this in steps, and unclear about "recent warm periods". *

## Russian impact zone

Will include review of the publication and the plan for evaluating scenarios of alternative fishing mortality in the Russian zone

## Ecosystem function idea

thinking on arguments for a EBS pollock "maintain ecosystem function" catch-advice rule as 
requested by the SSC. I think I mentioned something about management that provides a 
forage base (say of 1-3 yr old pollock) and avoids low levels. Carey and I chatted 
earlier (added her to this group) and realize that the estimates of these groups 
are relatively unreliable (i.e., 1-3 yr olds for next year are typically poorly estimated). 

Within the FMP we're talking about justification for catch advice below the maximum permissible. 
So what about the following semi-empirical approach:

  -  What if catch advice is adjusted up and down relative the historical mean biomass? I.e., if the catch 
  in the current year is say 1.2 million t, and the SSB next year is 30% above the mean, then with a 
  regulator to dampen change, next year’s recommendation would be 1.2 * (1.3)^0.5 =1.368 million t.      
  
  -  Similarly, if the SSB next year was only 75% of the mean value, the recommendation would be 1.2 * (0.75)^0.5 =1.039 million t.     
  
  -  If the SSB stayed at 75% of the mean, then the following year would be 1.039 * (.75)^0.5 = 0.8998 million t.          

The rationale proposed is that the detrimental effects of fishing on the ecosystem appear to be mainly sustainable.

Are there assignable issues from fishing affected the ecosystem function that we’ve measured with any degree of confidence?

If there hadn’t been any pollock fishing ever, would we be able to note that the ecosystem would be in a better state?
Tuning the adjustments based on historical catches and ecosystem outcomes would generally be deemed "acceptable"



# EBS 2023 assessment analyses planning

## From the 2022 SSC minutes:

**The SSC suggests that walleye pollock is a good candidate for
considering the impacts of highly variable recruitment on reference
points in the context of the Council's harvest control rules** (see
discussion on working groups in the JGPT report section). For example,
the SAFE authors suggested exploring an explicit harvest control rule
that maintains productivity at the level observed over recent decades
(p. 33). The SSC supports considerations of modified harvest control
rules, particularly for stocks with highly variable and uncertain
recruitment. If the Council chooses, this could include considerations
for stabilizing catches over time or including other economic
considerations in the harvest control rules.

-   *For the 2023 assessment we examine the variability of the biological reference points historically and note that there is general stability in the 
$B_{MSY}$ 
estimates. The Tier 1 ABC/OFL cthealculations can result in highly variable 
estimates as the stock approaches
$B_{MSY}$ and drops below that value (as happend in the 2009-2010 period).*


The SSC had the following additional recommendations for the authors:

-   Maturity and growth information from the NBS has not been examined
    yet. Given the possible importance of the NBS to walleye pollock and
    other species in the future, the SSC suggests this should be a high
    priority.

    -   *These data being processed and this work  is underway*

-   The SSC supports efforts to implement recent advances in improving
    the statistical treatment of compositional data using the Dirichlet
    distribution or other approaches.

    -   *Tradeoffs in data weighting were pursued in September 2023 and
        saught to find a balance between observation error and process
        errors*

-   The SAFE document lists a number of research recommendations (p.
    36/37). The SSC notes that some of these are at least in progress.
    The SSC generally supports these recommendations but requests that
    the authors update the list of priorities to clarify to what extent
    some of these priorities have been partially or fully addressed.

    -   *We updated the priorities and listed those that have been
        completed or are continued to be underway*"

-   In particular, the SSC notes that genetic sample collection and
    analyses are listed as a research priority across all pollock stocks
    and that some work has been completed. The SSC highlights the
    importance of additional genetics work and would appreciate an
    update on the status of this work either as part of the assessment
    or separately.

    -   *We revisited the stock structure work attached as an appendix to the 2015 
    SAFE report chapter and are examining the extent that this work
    needs updating.
        *"

-   The SSC appreciated the adjustments to weight-at-age in the survey
    that was included in this year's assessment and suggests that these
    changes may be substantial enough to warrant an examination of their
    impact on assessment results.

    -   *This publication has been completed and in the present assessment
    we evaluated the implication of alternative spawning biomass-at-age 
    assumptions.*"

-   With respect to the multi-species CEATTLE model, the SSC concurs
    with Plan Team recommendations to use the model to inform risk table
    discussions and to consider ways in which model outputs, in
    particular estimates of predation mortality, can inform
    single-species assessments.

    -   *We wil include a more thorough evaluation and discussion in the 
    2023 assessment.*

-   The SSC encourages the authors to consider model-based solutions to
    uncertain recruitment estimates rather than ad-hoc adjustments. In
    particular, reductions in the assumed recruitment variance parameter
    may result in less extreme recruitment estimates. Other systematic
    approaches to addressing the uncertainty may also be considered.

    -   *We revisited applying the age-determination error matrix as 
    a sensitivity as this can impact the recruitment variability and 
    estimation uncertainty.*

-   The SSC suggests that authors include a plot to compare estimates of
    recent recruitments as they change over time similar to Fig 3.33
    (pg. 88) in the sablefish assessment.

    -   *We provide a figure of estimated recruitment by year class
        (1977 -- 2019) in number of age-1 fish (billions of fish) for
        the 2022 and 2023 models.* "

-   The SSC supports the move across assessment from design-based
    estimates of survey biomass to VAST estimates. The SSC recommends
    that the design-based estimates be produced as a check on VAST
    estimates and as a fallback option if needed, although they may not
    need to be included in the assessment.

    -   *We provide a table showing the design-based estimates and conduct
    a model run with those estimates. This may be an approach to adopt so
    that bridging across assessment modeling platforms can be facilitated
    (most other assessment model platforms are unable to deal with index
    time series that have a covariance matrix)*

### Tasks and todo list

  [ ] Add in survey biomass\
  [ ] Update catches\
  [ ] Add in 2022 fishery age compositions\
  [ ] Add in 2023 survey age compositions\
  [x] Modify code to accommodate FT-NIRS and other age-determination error
  matrices by gear and year 
  [ ] Figure out which models to use going
    forward

## From the 2021 SSC minutes

-   Ongoing genetic studies to determine the relationship between
    pollock in the NBS and EBS, and nearby GOA and AI regions.

-   The 2019 BSAI GPT recommendation to revisit and evaluate the
    treatment of variance parameters within the assessment, with
    particular attention to those that are fixed.

-   Efforts to quantify pollock movement and abundance along the
    US-Russia EEZ boundary.

-   Geostatistical analyses of combined trawl and acoustic data to
    provide a single time-series, statistically accounting for the
    overlap between these data, for informing stock trends.

## The SSC provides the following additional recommendations:

-   Exploration of young-of-year pollock density and quality estimates
    from NMFS BASIS surveys to inform pollock recruitment.

-   Consideration of whether the observed sensitivity in the SRR to
    prior specification should constitute an increased risk level
    specification within the assessment or population dynamicsrelated
    considerations. This could provide a clearer justification for the
    use of the Tier 3 calculation as the basis for harvest
    specification.

-   Given the time-varying specification of fishery selectivity within
    the assessment model and the large change in the estimated 2021 FOFL
    between the 2019 and 2020 assessments, the authors should provide a
    retrospective comparison of the selectivity assumed in projections
    to that estimated with the addition of new data.

-   Consideration of whether risk table specifications should account
    for the importance of pollock as a key forage species in the EBS
    ecosystem to better justify the use of a Tier 3 ABC determination as
    a precautionary measure for this Tier 1 stock.

-   Given the apparent disappearance of the second and large mode in
    fishery length compositions as the 2020 B-season progressed,
    exploration of within-season spatial variation in fishery length
    composition would be useful in evaluating whether these larger
    pollock simply moved out of the area of fishing effort, or died as a
    result of natural or fishing mortality.
