---
title: "Application of one-step ahead residuals to alternative EBS pollock model specifications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example on how to change and run a model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo=FALSE,warning=FALSE,message=FALSE}
library(ebswp)
library(compResidual)
library(ggplot2)
library(cowplot)
library(reshape2)
library(dplyr)
library(tidyr)
theme_set(theme_bw())
library(here)
source(here::here("tools", "plot_osa_comps.R"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# One-step ahead (OSA) residuals

## EBS pollock models
We adopted the routine presented in @trijoulet2023 to evaluate residual
patterns in models. Below are the figures produced for the base model
from 2022 followed by the updated data and model configuration presented
in 2023.

#### Base model from 2022


```{r osa1,echo=FALSE,warning=FALSE,message=FALSE}
load(file="../doc/modtune.rdata")
ages <- 2:15
M<-(modtune[[1]])
o    <- M$pobs_ats[,3:16]
p    <- M$phat_ats[,3:16]
Neff <- M$EffN_ats[,2]
pearson <- Neff*(o-p)/sqrt(p*Neff)
years <- M$pobs_ats[,1]
plot_osa_comps(o,p, pearson, index=ages, years=years, index_label='age', Neff=Neff,
               stock='EBSpollock', survey='tuned_ats', outpath='doc/figs')
```


#### Updated model from September 2023

```{r osa2,echo=FALSE,warning=FALSE,message=FALSE}
M<-(modtune[[2]])
o    <- M$pobs_ats[,3:16]
p    <- M$phat_ats[,3:16]
Neff <- M$EffN_ats[,2]
pearson <- Neff*(o-p)/sqrt(p*Neff)
years <- M$pobs_ats[,1]
plot_osa_comps(o,p, pearson, index=ages, years=years, index_label='age', Neff=Neff,
               stock='EBSpollock', survey='tuned_ats', outpath='doc/figs')
```

# References

Trijoulet, Vanessa, Christoffer Moesgaard Albertsen, Kasper Kristensen, Christopher M. Legault, Timothy J. Miller, and Anders Nielsen. 2023. “Model Validation for Compositional Data in Stock Assessment Models: Calculating Residuals with Correct Properties.” Fisheries Research 257: 106487. https://doi.org/https://doi.org/10.1016/j.fishres.2022.106487.
