---
title: "SAFE report front matter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SAFE report front matter|
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ebswp)
thisyr=2022
nextyr=thisyr+1
lastyr=thisyr-1
```

# Introduction

## General

Walleye pollock (*Gadus chalcogrammus*; hereafter referred to as
pollock) are broadly distributed throughout the North Pacific with the
largest concentrations found in the Eastern Bering Sea. Also known as
Alaska pollock, this species continues to play important roles
ecologically and economically.
\clearpage

## Review of Life History

In the EBS pollock spawn generally in the period March-May and in
relatively localized regions during specific periods (Bailey 2000).
Generally spawning begins nearshore north of Unimak Island in March and
April and later near the Pribilof Islands (Jung et al. 2006, Bacheler et
al. 2010). Females are batch spawners with up to 10 batches of eggs per
female per year (during the peak spawning period). 
Eggs and larvae of EBS pollock are planktonic for a
period of about 90 days and appear to be sensitive to environmental
conditions. These conditions likely affect their dispersal into
favorable areas (for subsequent separation from predators) and also
affect general food requirements for over-wintering survival (Gann et
al. 2015, Heintz et al., 2013, Hunt et al. 2011, Ciannelli et al. 2004).
Duffy-Anderson et al. (2015) provide a review of the early life history
of EBS pollock.

Throughout their range juvenile pollock feed on a variety of planktonic
crustaceans, including calanoid copepods and euphausiids. In the EBS
shelf region, one-year-old pollock are found throughout the water
column, but also commonly occur in the NMFS bottom trawl survey. Ages 2
and 3 year old pollock are rarely caught in summer bottom trawl survey
gear and are more common in the midwater zone as detected by mid-water
acoustic trawl surveys. Younger pollock are generally found in the more
northern parts of the survey area and appear to move to the southeast as
they age (Buckley et al. 2009). Euphausiids, principally *Thysanoessa
inermis* and *T. raschii*, are among the most important prey items for
pollock in the Bering Sea (Livingston, 1991; Lang et al., 2000; Brodeur
et al., 2002; Cianelli et al., 2004; Lang et al., 2005). Pollock diets
become more piscivorous with age, and cannibalism has been commonly
observed in this region. However, Buckley et al. (2015) showed spatial
patterns of pollock foraging by size of predators. For example, the
northern part of the shelf region between the 100 and 200 m isobaths
(closest to the shelf break) tends to be more piscivorous than
counterparts in other areas.

## Stock structure

Stock structure for EBS pollock was evaluated in Ianelli
et al. (2015). In that review past work on genetics (e.g., Bailey et al.
1999, Canino et al. 2005) provided insight on genetic differentiation.
The investigation also compared synchrony in year-classes and growth
patterns. Pollock samples from areas including Zhemchug Canyon, Japan,
Prince William Sound, Bogoslof, Shelikof, and the Northern Bering Sea
were processed and results presented in 2021. 
This analysis included 617 walleye 
pollock from Japan, Bering Sea, Chukchi Sea, Aleutian Islands, Alaska 
Peninsula, and Gulf of Alaska. Results suggests there is temporally 
stable stock structure with a latitudinal gradient, i.e.,
 Bering Sea pollock are distinguishable from those in the
 Gulf of Alaska and Aleutian Islands (I. Spies, personal communication, 2021). 
 Notably, Bogoslof samples appeared genetically distinct from other spawning regions. 
 In addition, there appeared to be genetic connectivity between the eastern 
 Bering Sea and the western Gulf of Alaska spawning samples. Samples 
 from the eastern Gulf of Alaska are currently undergoing sequencing to 
 determine whether eastern Gulf of Alaska pollock are 
genetically distinct from those in the western Gulf of Alaska. 
The ongoing goals of the project include investigating the genetic stock structure of walleye pollock,
testing if patterns are temporally stable, and evaluating if
distributional shifts under climate change can be detected. 

For management purposes, the preliminary conclusions from these genetics
results are: 1)there is stock structure in pollock that appears to be
stable through time and 2) Some aspect of stock structure is
latitudinal---Bering Sea pollock appear distinct from fish collected
from the Gulf of Alaska and the Aleutian Islands. The results appear
strong enough that a GTseq panel could be designed in the future to
determine stock of origin of walleye pollock, the scale of which may be
relatively large, such as "Bering Sea" or "GOA". The scope and funding
sources for this project will be planned in 2022 with sampling designs
developed for implementation as early as 2023.
