---
title: "Chapter 1: Assessment of the Walleye Pollock Stock \\ in the Eastern Bering Sea"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    highlight: zenburn
    latex_engine: xelatex
    fig_caption: yes
    number_sections: false
    toc: false
    includes:
      in_header: "preamble.tex"
    keep_tex: yes
fontsize: 11pt
header-includes:
geometry: margin=1in
---
  
\pagenumbering{gobble}

```{r global_options, include=FALSE}
doplots       <<- FALSE
do_data_plots <<- FALSE
domcmc        <<- FALSE
doproj        <<- FALSE
source("../pm_21.R")
library(knitr)
opts_chunk$set(fig.width = 6, fig.height = 8, echo = FALSE, warning = FALSE, message = FALSE, cache=FALSE,results="asis",knitr.duplicate.label = "allow")
clean_cache(clean = TRUE)
#include = FALSE prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
#echo = FALSE prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
#message = FALSE prevents messages that are generated by code from appearing in the finished file.
#warning = FALSE prevents warnings that are generated by code from appearing in the finished.
#fig.cap = "..." adds a caption to graphical results.
#To set global options that apply to every chunk in your file, call knitr::opts_chunk$set in a code chunk. Knitr will treat each option that you pass to 
```

------------------------------------------------------------------------

```{=tex}
\begin{centering}
  James Ianelli, Ben Fissel, Sarah Stienessen, Taina Honkalehto, \\
  Elizabeth Siddon, and Caitlin Allen-Akselrud \\
----
    \\

  Alaska Fisheries Science Center, National Marine Fisheries Service \\
  National Oceanic and Atmospheric Administration \\ 
  7600 Sand Point Way NE., Seattle, WA 98115-6349 \\ `r format(Sys.time(), '%B %d, %Y')` \\ 
    \end{centering}
```

------------------------------------------------------------------------

```{r child = 'text.Rmd'}
```

```{r child = 'ref.Rmd'}
```

\pagebreak

```{r child = 'tab.Rmd'}
```

```{r child = 'figs.Rmd'}
```

```{r child = 'model.Rmd'}
```

\pagebreak

```{r child = 'App.Rmd'}
```

\pagebreak

# List of contents, tables and figures

\tableofcontents

\listoftables

\listoffigures
