---
output:
  pdf_document: default
  html_document: default
  output: 
  bookdown::pdf_document2:
    highlight: zenburn
    latex_engine: xelatex
    fig_caption: yes
    number_sections: false
    toc: false
    includes:
      in_header: "paramble.tex"
    keep_tex: yes
fontsize: 11pt
header-includes:
  
geometry: margin=1in
---
# Parameter estimates for the EBS pollock reference model 
Includes parameter set from Model 20.0c, uses VAST full time series (from bottome trawl survey) including Northern Bering Sea segment and the USV data for 2020 converted
to biomass (extending the ATS data) and with the stock-recruitment relationship that ignores the influence of the 1978 year class. This includes preliminary
size and age composition from the 2021 fishery.

\fontsize{7}{9}
```{r eval = TRUE, include=TRUE,echo=FALSE, comment=NA }
#opts_chunk$set(fig.width = 12, fig.height = 8, echo = FALSE, warning = FALSE, message = FALSE)
ts <- readLines("../runs/c21/pm.std")
for (i in 1:length(ts)) cat(ts[i], "\n")
```

# Main data file
\fontsize{3}{4}
```{r eval = TRUE, include=TRUE,echo=FALSE, comment=NA }
ts <- readLines("../runs/dat/pm_2021_w21catchage.dat")
for (i in 1:length(ts)) cat(ts[i], "\n")
```
