```{r summary}
#df <- dplyr::filter(data.frame(M[[2]]$dCatchData_out), V2 == 2, V3 %in% c(2,3)) %>% dplyr::select(V1, V3, V5) %>% tidyr::spread(key = V3, value =G V5) #%>%
 #dplyr::mutate(V1 = paste0(V1, "/", as.numeric(sub('.*(?=.{2}$)', '', V1, perl = TRUE)) + 1))
#names(df) <- c("Year","Trawl bycatch","Fixed gear bycatch")
#tab <- xtable(df, caption = "Groundfish SMBKC male bycatch biomass (tonnes) estimates. Trawl includes pelagic trawl and non-pelagic trawl types. Source: J. Zheng, ADF\\&G, and author estimates based on data from R. Foy, NMFS. AKRO estimates used after 2008/09.", label = "tab:pscbycatch", digits = c(0,0,3,3))
#print(tab, caption.placement = "top", include.rownames = FALSE)
npar <- 999
bmsy <- 2000 
b0   <- 5700 
b100 <- 6608 
b40  <- 2643 
b35  <- 2643 
# String versions of these values
bmsys  <- format(bmsy,big.mark=",",scientific=F)
b0s  <- format(b0,big.mark=",",scientific=F)
b100s  <- format(b100,big.mark=",",scientific=F)
b40s  <- format(b40,big.mark=",",scientific=F)
b35s  <- format(b35,big.mark=",",scientific=F)
thisyr <- 2017
nextyr <- thisyr+1
```
```{r junk}
print(M$T1)
```


# Executive summary

This chapter covers the Eastern Bering Sea (EBS) region—the Aleutian Islands
region (Chapter 1A) and the Bogoslof Island area (Chapter 1B) are presented
separately with full assessments expected in 2018. 

###Summary of changes in assessment inputs

Relative to last year’s BSAI SAFE report, the
following substantive changes have been made in the EBS pollock stock
assessment.

####Changes in the data

   1.	The 2017 NMFS bottom-trawl survey (BTS) biomass and abundance at age estimates were included. 

   2.	The 2016 NMFS acoustic-trawl survey (ATS) biomass and abundance at age estimates were updated based on
      age data collected from the ATS sampling (in 2016 the BTS age-length key was used). 
   2.  The time series from 1994-2016 of including the bottom 2.5 m from the acoustic data were completed and 
      available for including.
   2.  Two additional years of opportunistic acoustic data from vessels transiting the EBS shelf region 
     were processed and the time series now extends from 2006-2017. This provides an alternative
     index of pollock biomass in mid-water.
   2.	Observer data for catch-at-age and average weight-at-age from the 2016 fishery were finalized and included.
   2.	Total catch as reported by NMFS Alaska Regional office was updated and included through 2017.

###Changes in the assessment methods

Several modifications to the methods were adopted based on a review by the
Center for Independent Experts (CIE) and feedback from September/October 2016
presentations to the NPFMC’s Plan Team and SSC. This included changes to the
treatment of uncertainty in current-year fishery mean weights-at-age and those
used for near term projections. The surveys were fitted to biomass estimates
instead of abundance. Sample sizes specified for the robust-multinomial
likelihood were re-evaluated based on CIE comments and selectivity variability
examined. The method of estimating current and future year mean body weight at
age was updated (as presented in September/October of 2016) and used. An
alternative for specifying the stock-recruit relationship for projection
purposes was evaluated due to CIE concerns about the prior distribution as
applied. The latter increased the risk-averse buffer between ABC and OFL
(computed as 1-ABC/OFL) slightly (13\% to 14\%) with most of the change in value
arising from the lower estimate of stock-recruit relationship steepness
parameter.