
# Executive summary

This chapter covers the Eastern Bering Sea (EBS) region—the Aleutian Islands
region (Chapter 1A) and the Bogoslof Island area (Chapter 1B) are presented
separately with full assessments expected in 2018. 

###Summary of changes in assessment inputs

Relative to last year’s BSAI SAFE report, the
following substantive changes have been made in the EBS pollock stock
assessment.

####Changes in the data

   1.	The 2017 NMFS bottom-trawl survey (BTS) biomass and abundance at age estimates were included. 

   2.	The 2016 NMFS acoustic-trawl survey (ATS) biomass and abundance at age estimates were updated based on
      age data collected from the ATS sampling (in 2016 the BTS age-length key was used). 
   2.  The time series from 1994-2016 of including the bottom 2.5 m from the acoustic data were completed and 
      available for including.
   2.  Two additional years of opportunistic acoustic data from vessels transiting the EBS shelf region 
     were processed and the time series now extends from 2006-2017. This provides an alternative
     index of pollock biomass in mid-water.
   2.	Observer data for catch-at-age and average weight-at-age from the 2016 fishery were finalized and included.
   2.	Total catch as reported by NMFS Alaska Regional office was updated and included through 2017.

###Changes in the assessment methods

Several modifications to the methods were adopted based on a review by the
Center for Independent Experts (CIE) and feedback from September/October 2016
presentations to the NPFMC’s Plan Team and SSC. This included changes to the
treatment of uncertainty in current-year fishery mean weights-at-age and those
used for near term projections. The surveys were fitted to biomass estimates
instead of abundance. Sample sizes specified for the robust-multinomial
likelihood were re-evaluated based on CIE comments and selectivity variability
examined. The method of estimating current and future year mean body weight at
age was updated (as presented in September/October of 2016) and used. An
alternative for specifying the stock-recruit relationship for projection
purposes was evaluated due to CIE concerns about the prior distribution as
applied. The latter increased the risk-averse buffer between ABC and OFL
(computed as 1-ABC/OFL) slightly (13\% to 14\%) with most of the change in value
arising from the lower estimate of stock-recruit relationship steepness
parameter.


## Summary of EBS pollock results

\begin{table}[ht]
\centering
\label{tab:pscbycatch}
\begin{tabular}{lrrrr}
  \hline
       & \multicolumn{2}{c}{As estimated or $\mathit{specified}$ } & \multicolumn{2}{c}{As estimated or $\mathit{recommended}$ }  \\
       & \multicolumn{2}{c}{specified $\mathit{last}$ year for:}  & \multicolumn{2}{c}{$\mathit{this}$ year for: }               \\
        Quantity & `r thisyr`      &`r thisyr+1`   & `r thisyr+1`      &`r thisyr+2` \\ 
  \hline
	M (natural mortality rate, ages 3+)	&	0.3	&	0.3	&	0.3	&	0.3 \\
Tier	&	1a	&	1a	&	1a	&	1a \\
Projected total (age 3+) biomass (t)	&	13,000,000 t	&	12,100,000 t	&		&	 \\
Projected female spawning biomass (t) &	4,600,000 t	&	4,500,000 t	&		&	 \\
$B_0$                                 &	5,700,000 t	&	5,700,000 t	&		&	 \\
$B_{msy}$                             &	2,165,000 t	&	2,165,000 t	&		&	 \\
$F_{OFL}$                             &	0.465	&	0.465	&		&	 \\
$maxF_{ABC}$                        &	0.398	&	0.398	&		&	 \\
$F_{ABC}$ &	0.36	&	0.37	&		&	 \\
$OFL$ (t)	&	3,640,000 t	&	4,360,000 t	&		&	 \\
$maxABC$ (t)	&	3,120,000 t	&	3,740,000 t	&		&	 \\
$ABC$ (t)	&	2,800,000 t	&	2,979,000 t	&		&	 \\
Status	&	2015	&	2016	&		&	 \\
Overfishing	&	No	&	n/a	&	No	&	n/a \\
Overfished	&	n/a	&	No	&	n/a	&	No \\
Approaching overfished	&	n/a	&	No	&	n/a	&	No \\
\hline
\end{tabular}
\end{table}

## Response to SSC and Plan Team comments

### General comments
The Plan Teams noted that a compilation of responses to CIE reviews be
included in order to maximise their benefit and to promote transparency.

_A table summarizing key aspects from the three reviews conducted in 2016 and 
responses is provided in an attachment._

###Comments specific to this assessment

In the September 2016 minutes, the BSAI Plan Team recommended: \" .../ that the
authors develop a better prior for steepness, or at least a better rationale,
and perhaps consider a meta-analytic approach.  
...In the long term, the Team recommends evaluating the sample sizes used 
for the data weighting and pursuing other CIE suggestions.

_Sample size estimates were re-evaluated and used in the recommended model below. An
alternative degree of uncertainty, which notes differences from the CEATTLE
stock-recruit relationship was provided as an alternative._

 
#Introduction

##General

Walleye pollock (_Gadus chalcogrammus_; hereafter referred to as pollock) are
broadly distributed throughout the North Pacific with the largest
concentrations found in the Eastern Bering Sea. Also  known as
Alaska pollock, this species continues to play important roles ecologically
and economically.

##Review of Life History
In the EBS pollock spawn generally in the period March-May with different subareas
spawning earlier or later ().

Eggs and larvae of EBS pollock are planktonic for a period of about XX XX and
appear to be sensitive to environmental condations. These conditions likely affect
their dispersal into favorable areas (for subsequent separation from predators) and
also affect general food requirements and survival (Hintze, 20XX, Hunt et al. )

Throughout their range juvenile pollock feed on a variety of planktonic crustaceans, 
including calanoid copepods and euphausiids.

In the EBS shelf region, one-year-old pollock are found throughout the water column but
also commonly occur in the NMFS bottom trawl survey. Ages 2 and 3 year old pollock are
rarely caught in summer bottom trawl survey gear and are commonly occurring in the 
midwater zone as detected by mid-water acoustic trawl surveys. Younger pollock are 
generally found in the more northern parts of the survey area and a pattern of 
movement to the southeast occurs as the age (Buckley ). Their diets with age 
become more piscivorous and cannibalism has been commonly observed for this 
region. 


##Stock structure

New information available from ecosystem survey work in the Northern Bering
Sea (NBS) region (north of Nunivak Island to the Russian convention line and
into Norton Sound)  suggests considerably more pollock present there compared
to the 2010 survey (1.3 million t in 2017 compared to 11 thousand t in 2010).
Although the 2017 bottom temperatures were colder than recent years, the warm
conditions in 2016 may have caused a portion of the pollock stock to move into
this region. A loose relationship was determined (R^2 of 0.43) between mean
bottom temperature in the US zone on the EBS shelf and subsequent biomass
estimates in the  Navarin basin (the Russian area adjacent to the Convention
Line; Ianelli et al. 2011). However,  the extent that this may occur between
years is unknown and more detailed evaluation of the NBS data will be
forthcoming. Fortunately, genetics samples were taken from pollock and pending
funding availability, should help to ascertain the extent that these fish are
related to those observed in the normal EBS shelf survey area.   Genetics
samples taken from 2017 RACE summer survey from the Northern Bering Sea can be
compared with samples from the standard Bering Sea Unimak, Pribilof, and
Zhemchug, to ascertain the extent that these fish are related.


#Fishery 
##Description of the directed fishery

EBS pollock catches were low until directed foreign fisheries began in 1964.
Catches increased rapidly during the late 1960s and reached a peak in 1970-75
when they ranged from 1.3 to 1.9 million t annually (Fig. \ref{fig:catch}).
Following the peak catch in 1972, bilateral agreements with Japan and the USSR
resulted in reductions. Since 1977 (when the U.S. EEZ was declared) the annual
average EBS pollock catch has been about 1.2 million t, ranging from 0.815
million t in 2009 to nearly 1.5 million t during 2003-2006 (Fig.
\ref{fig:catch}). United States vessels began fishing for pollock in 1980 and
by 1987 they were able to take 99% of the quota. Since 1988, only U.S. vessels
have been operating in this fishery. Observers collected data aboard the
foreign vessels since the late 1970s. The current observer program for the
domestic fishery formally began in 1991 and has since then regularly re-
evaluated the sampling protocol and making adjustments where needed to improve
efficiency. Since 2011, regulations require that all vessels participating in
the pollock fishery carry at least one observer. Prior to this time about
70-80% of the catch was observed at sea or during dockside offloading. During
a 10-year period, catches by foreign vessels operating in the “Donut Hole”
region of the Aleutian Basin were substantial totaling nearly 7 million t
(Table \ref{tab:catch}).  A fishing moratorium was enacted in 1993 and only
trace amounts of pollock have been harvested from the Aleutian Basin region
since then.
<!-- XXXX and more text referencing Table \ref{`r paste0("tab:",tablab[1])`}   -->

###Catch patterns

The \“A-season\” for directed EBS pollock fishing opens on January 20th and
extends into early-mid April. During this season, the fishery produces highly
valued roe that, under optimal conditions, can comprise over 4% of the catch
in weight. The second, or “B-season” presently opens on June 10th and extends
through noon on November 1st. The A-season fishery concentrates primarily
north and west of Unimak Island depending on ice conditions and fish
distribution. There has also been effort along the 100 m contour (and deeper)
between Unimak Island and the Pribilof Islands. Since 2011, regulations and
industry-based measures to reduce salmon bycatch have affected the spatial
distribution of the fishery and to some degree, the way individual vessel
operators fish (Stram and Ianelli, 2014). The catch estimates by sex for the
A-season compared to estimates for the entire season indicate that over time,
the number of males and females has been fairly equal (Fig. \ref{fig:xxx}).
The 2016 and 2014 A-season fishery spatial pattern had relatively high
concentrations of fishing on the shelf north of Unimak Island, especially
compared to the pattern observed in 2015 when most fishing activity occurred
further north (Fig. \ref{fig:xxx}).

The 2017 summer and fall (B-season) fishing was had similar pattern to 2016
based on fleet-wide catch per hour fished (Fig. \ref{fig:`r figlab[4]`}
XXXCompared to 2011 B-season, the combined fleet took about one third of the
actual fishing time to reach 600 kt. Spatially, the 2016 B-season was much
more concentrated around the “horseshoe,” near the shelf break west of the
Pribilof Islands and extending north and west from Amak Island (Fig.
\ref{fig:xxx}). Since 1979 the catch of EBS pollock has averaged 1.19 million
t with the lowest catches occurring in 2009 and 2010 when the limits were set
to 0.81 million t due to stock declines (Table \ref{tab:catch_disc}) Pollock
retained and discarded catch (based on NMFS observer estimates) in the Eastern
Bering Sea and Aleutian Islands for 1991-2016 are shown in (Table
\ref{tab:catch_disc}). Since 1991, estimates of discarded pollock have ranged
from a high of 9.1% of total pollock catch in 1992 to recent lows of around
0.6%. These low values reflect the implementation of the Council’s Improved
Retention /Improved Utilization program. Prior to the implementation of the
AFA in 1999, higher discards may have occurred under the “race for fish” and
incidental catch of pollock that were below marketable sizes. Since
implementation of the AFA, the vessel operators have more time to pursue
optimal sizes of pollock for market since the quota is allocated to vessels
(via cooperative arrangements). In addition, several vessels have made gear
modifications to avoid retention of smaller pollock. In all cases, the
magnitude of discards counts as part of the total catch for management (to
ensure the TAC is not exceeded) and within the assessment. Bycatch of other
non-target, target, and prohibited species is presented in the section titled
Ecosystem Considerations below. In that section it is noted that the bycatch
of pollock in other target fisheries is more than double the bycatch of other
target species (e.g., Pacific cod) in the pollock fishery.


##Management measures

The EBS pollock stock is managed by NMFS regulations that provide limits on
seasonal catch. The NMFS observer program data provide near real-time
statistics during the season and vessels operate within well-defined limits.
TACs have commonly been set well below the ABC value and catches have usually
stayed within these constraints (Table \ref{tab:abc_tac}). Allocations of the TAC split
first with 10% to western Alaska communities as part of the Community
Development Quota (CDQ) program and the remainder between at-sea processors
and shore-based sectors. For characterization of CDQ program see Haynie (2014). 
Seung and Ianelli (2016) combined a fish population dynamics
model with an economic model to evaluate regional impacts.

Due to concerns over that groundfish fisheries may impact the population  
rebuilding of Steller sea lions, a number of management measures
have been implemented over the years. Some measures were designed to reduce the possibility
of competitive interactions between fisheries and Steller sea lions. For the
pollock fisheries, seasonal fishery catch and pollock biomass distributions
(from surveys) indicated that the apparent disproportionately high seasonal
harvest rates within Steller sea lion critical habitat could lead to reduced
sea lion prey densities. Consequently, management measures redistributed the
fishery both temporally and spatially according to pollock biomass
distributions. This was intended to disperse fishing so that localized harvest
rates were more consistent with annual exploitation rates. The measures
include establishing: 1) pollock fishery exclusion zones around sea lion
rookery or haulout sites; 2) phased-in reductions in the seasonal proportions
of TAC that can be taken from critical habitat; and 3) additional seasonal TAC
releases to disperse the fishery in time.

Prior to adoption of the above management measures, the pollock fishery
occurred in each of the three major NMFS management regions of the North
Pacific Ocean: the Aleutian Islands (1,001,780 km$^2$  inside the EEZ), the
Eastern Bering Sea (968,600 km$^2$), and the Gulf of Alaska (1,156,100 km$^2$). The
marine portion of Steller sea lion critical habitat in Alaska west of 150 $^{\circ}$ W
encompasses 386,770 km$^2$ of ocean surface, or 12\% of the fishery management
regions.

Prior to 1999, 84,100 km$^2$, or 22\% of critical habitat was closed to the
pollock fishery. Most of this closure consisted of the 10 and 20 nm radius
all-trawl fishery exclusion zones around sea lion rookeries (48,920 km$^2$, or
13% of critical habitat). The remainder was largely management area 518
(35,180 km$^2$, or 9% of critical habitat) that was closed pursuant to an
international agreement to protect spawning stocks of central Bering Sea
pollock.

In 1999, an additional 83,080 km$^2$ (21%) of critical habitat in the Aleutian
Islands was closed to pollock fishing along with 43,170 km$^2$  (11%) around sea
lion haulouts in the GOA and Eastern Bering Sea. In 1998, over 22,000 t of
pollock were caught in the Aleutian Island region, with over 17,000 t taken
within critical habitat region. Between 1999 and 2004 a directed fishery for
pollock was prohibited in this region. Subsequently, 210,350 km$^2$  (54%) of
critical habitat was closed to the pollock fishery. In 2000 the remaining
phased-in reductions in the proportions of seasonal TAC that could be caught
within the BSAI Steller sea lion Conservation Area (SCA) were implemented.

On the EBS shelf, an estimate (based on observer at-sea data) of the
proportion of pollock caught in the SCA has averaged about 38% annually.
During the A-season, the average is about 42% (in part because pre-spawning
pollock are more concentrated in this area during this period). The proportion
of pollock caught within the SCA varies considerably, presumably due to
temperature regimes and population age structure. The annual proportion of
catch within the SCA varies and has ranged from an annual low of 11% in 2010
to high of 51% in 2016 (Table \ref{tab:sca_catch}). This high value was due to B-season
conditions which had 62% of the catch taken in this region.

The 1998 American Fisheries Act (AFA) reduced the capacity of the
catcher/processor fleet and permitted the formation of cooperatives in each
industry sector by the year 2000. Because of some of its provisions, the AFA
gave the industry the ability to respond efficiently to changes mandated for
sea lion conservation and salmon bycatch measures. Without such a catch-share
program, these additional measures would likely have been less effective and
less economical (Strong and Criddle 2014).

An additional strategy to minimize potential adverse effects on sea lion
populations is to disperse the fishery throughout more of the pollock range on
the Eastern Bering Sea shelf. While the distribution of fishing during the
A-season is limited due to ice and weather conditions, there appears to be
some dispersion to the northwest area (Fig. \ref{fig:catch_dist}).

Based on analysis, the majority (~56%) of Chinook salmon caught as bycatch in
the pollock fishery originate from western Alaskan rivers. An Environmental
Impact Statement (EIS) was completed in 2009 in conjunction with the Council’s
recommended management approach. This EIS evaluated the relative impacts of
different bycatch management approaches as well as estimated the impact of
bycatch levels on adult equivalent salmon (AEQ) returning to river systems
(NMFS/NPFMC 2009). As a result, revised salmon bycatch management measures
went into effect in 2011 which imposed new prohibited species catch (PSC)
limits. These limits, when reached,  close the fishery by sector and season
(Amendment 91 to the Groundfish FMP resulting from the NPFMC’s 2009 action).
Previously, all measures for salmon bycatch imposed seasonal _area_ closures
when PSC levels reached the limit (fishing could continue outside of the
closed areas). The current program imposes a dual cap system by fishing sector
and season. A  goal of this system was to maintain incentives to avoid bycatch
at a broad range of relative salmon abundance. Participants are also required
to take part in an incentive  program agreement (IPA). These IPAs are approved
and reviewed annually by NMFS  to ensure individual vessel accountability. The
fishery has been operating under  rules to implement this program since
January 2011.

Further measures to reduce salmon bycatch in the pollock fishery were
developed and the Council took action on Amendment 110 to the BSAI Groundfish
FMP in April 2015. These additional measures were designed to add protection
for Chinook salmon by imposing more restrictive PSC limits in times of low
western Alaskan Chinook salmon abundance. This included provisions within the
IPAs that reduce fishing in months of higher bycatch encounters and mandate
the use of salmon excluders in trawl nets. These provisions were also included
to manage chum salmon bycatch within the IPAs rather than through Amendment 84
to the FMP. The new measure also included additional seasonal flexibility in
pollock fishing so that more pollock (proportionally) could be caught during
seasons when salmon bycatch rates were low. Specifically, an additional 5% of
the pollock can be caught in the Aseason (effectively changing the seasonal
allocation from 40% to 45%, and this has occurred in 2017 (Fig.
\ref{fig:prop_a_season})). These measures are all part of Amendment 110 and a
summary of this and other key management measures is provided in Table
\ref{tab:mgt}.



##Economic conditions as of 2016

Alaska pollock is the dominant species in terms of catch in the Bering Sea \&
Aleutian Island (BSAI) region. It accounted for 69\% of the BSAI's FMP
groundfish harvest and 88\% of the total pollock harvest in Alaska. Retained
catch of pollock increased 2.4\% to 1.35 million t in 2016. BSAI pollock
first-wholesale value was \$1.35 billion 2016, which was up from \$1.27 billion
in 2015 and above the 2005-2007 average of \$1.25 billion. The higher revenue in
recent years is largely the result of increased catch and production levels as
the average first-wholesale price of pollock products have declined since
peaking in 2008-2010 and since 2013 have been close the 2005-2007 average,
though this varies across products types.

Pollock is targeted exclusively with pelagic trawl gear. The catch of pollock
in the BSAI was rationalized with the passage of the American Fisheries Act
(AFA) in 1998,\footnote{The AFA was implemented in 1999 for
catcher/processors, and in   2000 for catcher vessel and motherships.} which,
among other things, established a proportional allocation of the total
allowable catch (TAC) among vessels in sectors which were allowed to form into
cooperatives.\footnote{The   BSAI pollock TAC is divided between Community
Development Program (10\% off   the top), catcher vessels (50\%),
catcher/processors (40\%) and motherships   (10\%).}  Alaska caught pollock in
the BSAI became certified by the Marine Stewardship Council (MSC) in 2005, a
NGO based third-party sustainability certification, which some buyers seek. In
2015 the official U.S. market name changed from ``Alaska pollock'' to
``pollock'' enabling U.S. retailers to differentiate between pollock caught in
Alaska and Russia.

Prior to 2008 pollock catches were high at approximately 1.4 million t in the
BSAI for an extended period (Table \ref{tab:bsaiplck_exv}). The U.S. accounted
for over 50\% of the global pollock catch (Table~\ref{tab:bsaiplck_wsl}).
Between 2008-2010 conservation reductions in the pollock total allowable catch
(TAC) trimmed catches to an average 867 thousand t. The supply reduction
resulted in price increases for most pollock products, which mitigated the
short-term revenue loss (Table~\ref{tab:bsaiplck_glbl}). Over this same
period, the pollock catch in Russia increased from an average of 1 million t
in 2005-2007 to 1.4 million t in 2008-2010 and Russia's share of global catch
increased to over 50\% and the U.S. share decreased to 35\%. Russia lacks the
primary processing capacity of the U.S. and much of their catch is exported to
China and is re-processed as twice-frozen fillets. Around the mid- to late-
2000s, buyers in Europe, an important segment of the fillet market, started to
source fish products with the MSC sustainability certification, and some major
retailer in the U.S. later began to follow suit. Asian markets, an important
export destination for a number of pollock products, have shown less interest
in requiring MSC certification. The U.S. was the only producer of MSC
certified pollock until 2013 when roughly 50\% of the Russian catch became MSC
certified. Since 2010 the U.S. pollock stock rebounded with catches in the
BSAI ranging from 1.2-1.3 million t and Russia's catch has stabilized at 1.5
to 1.6 million t. The majority of pollock is exported; consequently exchange
rates can have a significant impact on market dynamics, particularly the
Dollar-Yen and Dollar-Euro.\footnote{Aggregate exports in
Table~\ref{tab:bsaiplck_glbl} may not fully account for   all pollock exports
as products such as meal, minced fish and other ancillary   product may be
coded as generic fish type for export purposes.} Additionally, pollock more
broadly competes with other whitefish that, to varying degrees, can serve as
substitutes depending on the product.

This market environment accounts for some of the major trends in prices and
production across product types. Fillet prices peaked in 2008-2010 but
declined afterwards because of the greater supply from U.S. and Russia. The
2013 MSC certification of Russian-caught pollock enabled access to segments of
European and U.S. fillet markets, which has put continued downward pressure on
prices. Pollock roe prices and production have declined steadily over the last
decade as international demand has waned with changing consumer preferences in
Asia. Additionally, the supply of pollock roe from Russia has increased with
catch. The net effect has been not only a reduction in the supply of roe from
the U.S. industry, but also a significant reduction in roe prices which are
roughly half pre-2008 levels. Prior to 2008, roe comprised 23\% of the U.S.
wholesale value share, and since 2011 it has been roughly 10\%. With the U.S.
supply reduction in 2008-2010, surimi production from pollock came under
increased pressure as U.S. pollock prices rose and markets sought cheaper
sources of raw materials (see Guenneugues and Ianelli 2013 for a global review
of surimi resources and market). This contributed to a growth in surimi from
warm- water fish of southeast Asia. Surimi prices spiked in 2008-2010 and have
since tapered off as production from warm-water species increased (as has
pollock). A relatively small fraction of pollock caught in Russian waters is
processed as surimi. Surimi is consumed globally, but Asian markets dominate
the demand for surimi and demand has remained strong.

The catch of pollock can be broadly divided between the shore-based sector
where catcher vessels make deliveries to inshore processors, and the at-sea
sector where catch is processed at-sea by catcher/processors and motherships
before going directly to the wholesale markets. The retained catch of the
shore-based sector increased 2.5\% increase to 704 thousand t. The value of
these deliveries (shore-based ex-vessel value) totaled \$209.4 million in
2016, which was down 7.9\% from the ex-vessel value in 2015, as the increased
catch was offset by a 9.7\% decrease in the ex-vessel price
(\ref{tab:bsaiplck_exv}). The first-wholesale value of pollock products was
\$808 million for the at-sea sector and \$543 million for the shore-based
sector (Table~\ref{tab:bsaiplck_wsl}). The higher revenue in recent years is
largely the result of increased catch levels as the average price of pollock
products have declined since peaking in 2008-2010 and since 2013 have been
close to the 2005-2007 average, though this varies across products types. The
average price of pollock products in 2016 increased for the at-sea sector and
shore-based sectors, which was largely attributable to an increase in the
price of roe products, though prices increased for fillets and surimi products
as well.

The portfolios of products shore-based and at-sea processors produce are
similar. In both sectors the primary products processed from pollock are
fillets, surimi and roe, with each accounting for approximately 40\%, 35\%,
and 10\% of first-wholesale value (Table~\ref{tab:bsaiplck_wsl}). The price of
products produced at-sea tend to be higher than comparable products produced
shore-based because of the shorter time span between catch, processing and
freezing. The price of fillets produced at-sea tend to be about 6\% higher,
surimi prices tend to be about 20\% higher and the price of roe about 45\%
higher. Average prices for fillets produced at-sea also tend to be higher
because they produce proportionally more higher-priced fillet types (like
deep-skin fillets). The at-sea price first wholesale premium averaged roughly
\$0.30 per pound between 2005-2010 but has decreased to an average of \$0.20
per pound since 2011, in part, because the shore-based sector increased their
relative share of surimi production.\footnote{The at-sea price premium is the
difference between the   average price of first-wholesale products at-sea and
the average price of   first-wholesale products shore-based.}

###Pollock fillets

A variety of different fillets are produced from pollock, with pin-bone-out
(PBO) and deep-skin fillets accounting for approximately 70\% and 30\% of
production in the BSAI, respectively. Total fillet production decreased 3.4\%
to 161 kt in 2016, but since 2010 has increased with aggregate
production and catch and has been higher than the 2005-2007 average
(Table~\ref{tab:bsaiplck_wsl}). The average price of fillet products in the
BSAI increased 4\% to \$1.41 per pound and is below the inflation adjusted
average price of fillets in 2005-2007 of \$1.46 per pound. Media reports
indicate that H\&G and fillet prices tended to be low throughout the year. The
small size of fish in the catch, significant inventories, and insolvency of a
major international pollock trader were cited as contributing factors. Low
H\&G prices incentivize Russia producers upgrade their fillet production
capacity in the near future, though fillets still a small portion of their
primary production,. Much of the Russian catch already goes to China for
secondary processing into fillets so this would do little to increase the
overall volume, however, increased primary fillet processing in Russia could
increase competition with U.S. produced single-frozen fillet products.
Approximately 30\% of the fillets produced in Alaska are estimated to remain
in the domestic market, which accounts for roughly 45\% of domestic pollock
fillet consumption.\footnote{Additionally, roughly 10\% of the at-sea   BSAI
production is processed as H\&G which is mostly exported, primarily to
China, where is reprocessed as fillets and some share of which returns to the
U.S.. China also processes H\&G from Russia into fillets which are also
imported into the domestic market. Current data collection does not allow us
to estimate the share of U.S. returning imports.} As recent fillet markets
have become increasingly tight, the industry has tried to maintain value by
increasing domestic marketing for fillet based product and creating product
types that are better suited to the American palette, in addition to increased
utilization of by-products.

###Surimi seafood

Surimi production continued an increasing trend through 2016, but at a more
moderate rate of 1.6\% to 190.8 thousand t which is above the 2005-2007
average. Prices have increased since 2013 to \$1.19 per pound in the BSAI in
2016 (Table~\ref{tab:bsaiplck_wsl}). Because surimi and fillets are both made
from pollock meat, activity in the fillet market can influence the decision of
processors to produce surimi. Industry news indicated the average size of fish
caught is down, which incentivizes surimi production because it yields a
higher value than fillets. Additionally, the supply of raw surimi material
continues to be constrained in Japan. The high volume of surimi production has
raised concerns that prices may begin to plateau or fall, but the more
favorable exchange rate with Japan in 2016 may have helped to shore up prices.

###Pollock roe

Roe is a high priced product that is the focus of the A season catch destined
primarily for Asian markets. Roe production in the BSAI tapered off in the
late-2000s and since has generally fluctuated at under 20 thousand t annually,
production averaged 27 thousand t in 2005-2007 and was 14.3 thousand t in 2016,
which is 24\% below production in 2015 (Fig. 1.6). Prices peaked in the
mid-2000s prices have followed a decreasing trend over the last decade which
continued until 2015. In 2016 roe production from the U.S. and Russia were low
as a result of the average size of fish caught being down, which also reduced
average grade of roe sold. Lower production and tight inventories put upward
pressure on roe prices. Additionally, the Yen to U.S. Dollar exchange rate was
more favorable in the 2016 than 2015. The net result in the BSAI was a 24\%
price increase in 2016 \$2.84 per pound, and value was down only 6\% to \$89
million (Table~\ref{tab:bsaiplck_wsl}).

###Fish oil

Using oil production per 100 tons as a basic index (tons of oil per ton
retained catch) shows increases for the at-sea sector. In 2005-2007 it was
0.3\% and starting in 2008 it increased and leveled off after 2010 with  over
1.5\% of the catch being converted to fish oil (Table~\ref{tab:bsaiplck_oil}).
This represents about a 5-fold increase in recorded oil production during this
period. Oil production from the shore-based fleet was somewhat higher than the
at-sea processors prior to 2008 but has been relatively stable according to
available records. Oil production estimates from the shore-based fleet may be
biased low because some production occurs at secondary processors (fishmeal
plants) in Alaska. The increased production of oil beginning in 2008 can be
attributed to the steady trend to add more value per ton of fish landed.

#Data

The following lists the data used in this assessment: 

| Source | Type | Years |
|----------------------------|-----------------------|----------------------------------------------|
| Fishery | Catch biomass | 1964-2017 |
| Fishery | Catch age composition | 1964-2016 |
| Fishery | Japanese trawl CPUE | 1965-1976 |
| EBS bottom trawl | Area-swept biomass and age-specific proportions | 1982-2017 |
| Acoustic trawl survey | Biomass index and age-specific proportions | 1994, 1996, 1997, 1999, 2000, 2002, 2004, 2006-2010, 2012, 2014, 2016 |
| Acoustic vessels of opportunity (AVO) | Biomass index | 2006-2017 |

##Fishery 
###Catch
The catch-at-age composition was estimated using the methods described by
Kimura (1989) and modified by Dorn (1992). Length-stratified age data are used
to construct age-length keys for each stratum and sex. These keys are then
applied to randomly sampled catch length frequency data. The stratum-specific
age composition estimates are then weighted by the catch within each stratum
to arrive at an overall age composition for each year. Data were collected
through shore-side sampling and at-sea observers. The three strata for the EBS
were:  i) January–June (all areas, but mainly east of 170°W); ii) INPFC area
51 (east of 170°W) from July–December; and iii) INPFC area 52 (west of 170°W)
from July–December. This method was used to derive the age compositions from
1991-2015 (the period for which all the necessary information is readily
available). Prior to 1991, we used the same catch-at-age composition estimates
as presented in Wespestad et al. (1996).

The catch-at-age estimation method uses a two-stage bootstrap re-sampling of
the data. Observed tows were first selected with replacement, followed by re-
sampling actual lengths and age specimens given that set of tows. This method
allows an objective way to specify the effective sample size for fitting
fishery age composition data within the assessment model. In addition,
estimates of stratum-specific fishery mean weights-at-age (and variances) are
provided which are useful for evaluating general patterns in growth and growth
variability. For example, Ianelli et al. (2007) showed that seasonal aspects
of pollock condition factor could affect estimates of mean weight-at-age. They
showed that within a year, the condition factor for pollock varies by more
than 15%, with the heaviest pollock caught late in the year from October-
December (although most fishing occurs during other times of the year) and the
thinnest fish at length tending to occur in late winter. They also showed that
spatial patterns in the fishery affect mean weights, particularly when the
fishery is shifted more towards the northwest where pollock tend to be smaller
at age. In 2011 the winter fishery catch consisted primarily of age 5 pollock
(the 2006 year class) and later in that year age 3 pollock (the 2008 year
class) were present. In 2012 - 2015 the 2008 year class been prominent in the
catches with 2015 showing the first signs of the 2012 year-class as three
year-olds in the catch (Fig. \ref{fig:xxx}; Table \ref{tab:fsh_age}). The sampling
effort for age determinations and lengths is shown in Tables 1.11 and 1.12.
Sampling for pollock lengths and ages by area has been shown to be relatively
proportional to catches (e.g., Fig. 1.8 in Ianelli et al. 2004). As part of
the re-evaluation of sample sizes assumed within the assessment, the number of
ages and lengths (and number of hauls from which samples were collected) show
significant changes over time (Fig. \ref{fig:xxx}). This information was used to inform
periods from which input sample size re-weighting was appropriate for
modeling. Regarding the precision of total pollock catch biomass, Miller
(2005) estimated the CV to be on the order of 1%.

Scientific research catches are reported to fulfill requirements of the
Magnuson-Stevens Fisheries Conservation and Management Act. The annual
estimated research catches (1963 - 2015) from NMFS surveys in the Bering Sea
and Aleutian Islands Region are given in (Table \ref{tab:res_catch}). Since
these values represent extremely small fractions of the total removals
(~0.02%) they are ignored as a contributor to the catches as modeled for
assessment purposes.

```{r child = 'survtext.Rmd'}
```

```{r child = 'restext.Rmd'}
```

##ABC Recommendation

ABC levels are affected by estimates of $F_{MSY}$ which depends principally on 
   1.  the estimated stock-recruitmentXXX, 
   2.  demographic schedules such as selectivity-at-age, maturity, growth), 
   3.  the BMSY level, and current stock size (both
spawning and fishable). Updated data and analysis result in an estimate of
2016 spawning biomass (4,070 kt) that is about 212% of BMSY (2,165 kt). The
replacement yield—defined as the catch next year that is expected to achieve a
2018 spawning biomass estimate equal to that from 2016—is estimated to be
about 2,500,000 t.

The EBS pollock stock appears to have rebounded from the 2008 low point and
shows significant increases due to two strong year classes (2008 and 2012).
However, there remain several concerns about the medium-term stock conditions.
Namely,

1.  The conditions in summer 2017 followed a warm period, precaution may be
warranted since warm conditions are thought to negatively affect the survival
of larval and juvenile pollock.

2.  The near-term prognosis for  survey found very few one-year-old pollock in
summer 2016 and 2017 (the BTS data show below average 1-year olds).

3.  The current BTS data show low abundances of pollock aged 10 and older.
Historically there had been good representation of older fish in data from
this survey. This is somewhat expected given the poor year-classes observed
during the period 2000-2005.

4.  The BTS showed patchier concentrations of pollock compared to recent
years. This can result in increased uncertainty in the estimates. This
patchier distribution may also reflect somewhat better nominal fishery catch
rates.

5.  The multispecies model suggests that the BMSY level is around 3.6 million
t instead of the ~2 million t estimated in the current assessment (noting that
the total natural mortality is higher in the multispecies model).

6.  Roe production has dropped in 2015 in the B-season. Recent data show that
~15% of annual roe production has occurred from June-October whereas in 2015
and 2016 the production is ~5%.

7.  The selection of a single model, though attempting to account for
uncertainties due to process errors, ignores structural uncertainty in model
specification. Including such structural uncertainties may reflect the type of
variability in stock-recruit relationship depicted in the scenario where
conditioning the curve on the assessment results is lowered.

8.  The euphausiid index (see Ecosystem considerations, this volume) decreased
from the 2014 estimates and has declined since the 2009 peak. This may
negatively affect survival rates of juvenile pollock prior to recruiting to
the fishery.

9.  Pollock are an important prey species for the ecosystem; there’s been a
12% decline in St. Paul Island pup production from 2014-2016 which, when
combined information on the other fur seal population components (Bogoslof and
St. George Islands), indicates an estimated 2.5% decline in the overall
Eastern Stock fur seal population.  Maintaining prey availability may provide
better foraging opportunities for the fur seal stock to minimize further
declines.

10. Whilst outside of ABC considerations, it seems that maintaining the stock
at relatively high levels and achieving fishery catch rates observed in 2016
B-season may help to minimize Chinook salmon bycatch (noting that the total
effort required to catch 600 kt in the 5 most recent B-seasons was
substantially smaller this year)

Given these factors, a `r nextyr` ABC of 2,800,000 t is recommended based on the
Tier 3 estimates as conservatively selected by the SSC in 2014 and 2015. We
recognize that the actual catch will be constrained by other factors (the 2
million t OY BSAI groundfish catch limit; bycatch avoidance measures). The
alternative maximum permissible Tier 1a ABC seems clearly risky. Such high
catches would result in unprecedented variability and removals from the stock
(and considerably more capacity and effort). Adopting a more stable catch
system would also result in less spawning stock variability.

#Ecosystem considerations
In general, a number of key issues for ecosystem conservation and management can be highlighted. These include: 

*	Preventing overfishing;    

*	Avoiding habitat degradation;    

*	Minimizing incidental bycatch;    

*	Monitoring bycatch and the level of discards; and    

*	Considering multi-species trophic interactions relative to harvest policies.    

For the case of pollock in the Eastern Bering Sea, the NPFMC and NMFS continue
to manage the fishery on the basis of these issues in addition to the single-
species harvest approach (Hollowed et al. 2011). The prevention of overfishing
is clearly set out as the main guideline for management. Habitat degradation
has been minimized in the pollock fishery by converting the industry to
pelagic-gear only. Bycatch in the pollock fleet is closely monitored by the
NMFS observer program and managed on that basis. Discard rates of many species
have been reduced in this fishery and efforts to minimize bycatch continue.

In comparisons of the Western Bering Sea (WBS) with the Eastern Bering Sea
using mass-balance food-web models based on 1980-85 summer diet data, Aydin et
al. (2002) found that the production in these two systems is quite different.
On a per-unit-area measure, the western Bering Sea has higher productivity
than the EBS. Also, the pathways of this productivity are different with much
of the energy flowing through epifaunal species (e.g., sea urchins and
brittlestars) in the WBS whereas for the EBS, crab and flatfish species play a
similar role. In both regions, the keystone species in 1980-85 were pollock
and Pacific cod. This study showed that the food web estimated for the EBS
ecosystem appears to be relatively mature due to the large number of
interconnections among species. In a more recent study based on 1990-93 diet
data (see Appendix 1 of the Ecosystem Considerations chapter for methods),
pollock remain in a central role in the ecosystem. The diet of pollock is
similar between adults and juveniles with the exception that adults become
more piscivorous (with consumption of pollock by adult pollock representing
their third largest prey item).

Regarding specific small-scale ecosystems of the EBS, Ciannelli et al. (2004a,
2004b) presented an application of an ecosystem model scaled to data available
around the Pribilof Islands region. They applied bioenergetics and foraging
theory to characterize the spatial extent of this ecosystem. They compared
energy balance, from a food web model relevant to the foraging range of
northern fur seals and found that a range of 100 nautical mile radius encloses
the area of highest energy balance representing about 50% of the observed
foraging range for lactating fur seals. This has led to a hypothesis that fur
seals depend on areas outside the energetic balance region. This study
develops a method for evaluating the shape and extent of a key ecosystem in
the EBS (i.e., the Pribilof Islands). Furthermore, the overlap of the pollock
fishery and northern fur seal foraging habitat (see Sterling and Ream 2004,
Zeppelin and Ream 2006) will require careful monitoring and evaluation.

A brief summary of these two perspectives (ecosystem effects on pollock stock
and pollock fishery effects on ecosystem) is given in (Table \ref{tab:pscbycatch}). Unlike the
food-web models discussed above, examining predators and prey in isolation may
overly simplify relationships. This table serves to highlight the main
connections and the status of our understanding or lack thereof.


## Ecosystem effects on the EBS pollock stock 

The pollock stock condition appears to have benefitted substantially from the
recent conditions in the EBS. The conditions on the shelf during 2008
apparently affected conditions for age-0 northern rock sole due to cold
conditions and apparently unfavorable currents that retain them into the over-
summer nursery areas (Cooper et al. 2014). It may be that such conditions
favor pollock recruitment. Hollowed et al. (2012) provided an extensive review
of habitat and density for age-0 and age-1 pollock based on extensive survey
data. They noted that during cold years, age-0 pollock were distributed
primarily in the outer domain in waters greater than 1ºC and during warm
years, age-0 pollock were distributed mostly in the middle domain. This
temperature relationship, along with interactions with available food in
early-life stages, appears to have important implications for pollock
recruitment success (Coyle et al. 2011). The fact that the 2012 year-class,
while uncertain, appears to be also high creating a favorable stock trend in
the near term.

A separate section presented this year updates multispecies model with more
recent data and is presented as a supplement to the BSAI SAFE report. In this
approach, a number of simplifications for the individual species data and
fisheries processes (e.g., no time varying selectivity in the fishery and only
design-based survey indices). However, that model mimics the pattern and
abundances with the single species reasonably well. It also allows specific
questions to be addressed regarding pollock TACs. For example, since predation
(and cannibalism) is explicitly modeled, the impact of relative stock sizes on
subsequent recruitment to the fishery can be now be directly estimated and
evaluated (in the model presented here, cannibalism is explicitly accounted
for in the assumed Ricker stock-recruit relationship).

Euphausiids, principally Thysanoessa inermis and T. raschii, are among the
most important prey items for pollock in the Bering Sea (Livingston, 1991;
Lang et al., 2000; Brodeur et al., 2002; Cianelli et al., 2004; Lang et al.,
2005). Buckley et al. (2016) showed spatial patterns of pollock foraging by
size of predators. For example, the northern part of the outer domain (closest
to the shelf break) tends to be more piscivorous than counterparts in other
areas (Fig. \ref{fig:xxx}). This figure also shows that euphausiids make up a larger
component of the diet in the southern areas. The euphausiid abundance on the
Bering Sea shelf is presented as a section of the 2016 Ecosystem
Considerations Chapter of the SAFE report and shows a continued decline in
abudance since the peak in 2009 (for details see De Robertis et al. (2010) and
Ressler et al. (2012). The role that the apparent recent 2009 peak abundance
had in the survival of the 2008 year class of EBS pollock is interesting.
Contrasting this with how the feeding ecology of the 2012 year class (also
apparently well above average) may differ is something to evaluate in the
future.

##EBS pollock fishery effects on the ecosystem. 

Since the pollock fishery is primarily pelagic in nature, the bycatch of non-
target species is small relative to the magnitude of the fishery (Table \ref{tab:pscbycatch}).
Jellyfish represent the largest component of the bycatch of non-target species
and had averaged around 5-6 thousand tons per year but more than doubled in
2014 but has dropped in 2015. The data on non-target species shows a high
degree of inter-annual variability, which reflects the spatial variability of
the fishery and high observation error. This variability may reduce the
ability to detect significant trends for bycatch species.

The catch of other target species in the pollock fishery represent less than
1% of the total pollock catch. Incidental catch of Pacific cod has increased
since 1999 but remains below the 1997 levels (Table \ref{tab:pscbycatch}). The incidental
catch of flatfish was variable over time and has increased, particularly for
yellowfin sole. Proportionately, the incidental catch has decreased since the
overall levels of pollock catch have increased. In fact, the bycatch of
pollock in other target fisheries is more than double the bycatch of target
species in the pollock fishery (Table \ref{tab:pscbycatch}).

The number of non-Chinook salmon (nearly all made up of chum salmon) taken incidentally
has steadily increased since 2014 with 2017 number in excess of 465 thousand fish (more than
double the 2003-2017 average of 227 thousand fish; Table \ref{tab:pscbycatch}).
Chinook salmon bycatch has also increased steadily since 2012 with the 2017
counts at just over 30,000 (which is 18% below the 2003-2017 mean value). 
However, this is the highest value since
the implementation of Amendment 91 in 2011 (Table \ref{tab:pscbycatch}). 
Ianelli and Stram (2014) provided estimates of the bycatch impact on Chinook salmon runs to the coastal west
Alaska region and found that the peak bycatch levels exceeded 7% of the total
run return. Since 2011, the impact has been estimated to be below 2%. Updated estimates 
given new genetic information and these levels of PSC will be provided in the future. 


#Data gaps and research priorities

The available data for EBS pollock are extensive yet many processes behind the
observed patterns are poorly understood. For example, the recent bottom trawl
surveys found abundance levels for the 2008  and now 2012 year class appear to
be estimated at high levels. Research on developing and testing plausible
hypotheses about the underlying processes that cause such observations is
needed. This should include examining potential effects of temporal changes in
survey stations and using spatial processes for estimation purposes (e.g.,
combining acoustic and bottom trawl survey data). The application of the
geostatistical methods (presented for comparative purposes above) seem like a
reasonable approach to statistically model disparate data sources for
generating better abundance indices.

More studies on spatial dynamics, including the relationship between climate
and recruitment and trophic interactions of pollock within the ecosystem would
be useful for improving ways to evaluate the current and alternative fishery
management system. In particular, studies investigating the processes
affecting recruitment of pollock in the different regions of the EBS
(including potential for influx from the GOA) should be pursued.

Many studies have found inconclusive evidence for genetic population structure
in walleye pollock. Knowledge of stock structure is particularly important for
this species, given its commercial importance. Therefore, a large scale study
using the highest resolution genetic tools available is recommended. Such a
study would incorporate samples throughout the range of walleye pollock,
including North America, Japan, and Russia, if possible. Data from thousands
of SNP loci should be screened, using next generation sequencing.

#Acknowledgements

We thank the staff of the AFSC age-and-growth department for their continued
excellence in promptly processing the samples used in this assessment. The
work of many individuals involved in collecting and processing survey and
observer data is greatly appreciated. Finally, thanks to the many colleagues who provided
edits and suggestions to improve this document.

